Worflow

### 2017-05-31 R/workflow.Rmd
update code for new data and add makefile

```{r}
runZinb = FALSE
runClus = FALSE
NCORES = 2
```

@@

@@

```{r datain}
load("../data/Expt4c2b_filtdata.Rda")
load("../data/E4c2b_slingshot_wsforkelly.RData")
```

Here we only look at the 1000 most variable genes.
```{r filtering}
names(batch) <- colnames(counts)

counts <- counts[,names(clus.labels)]
batch <- droplevels(batch[names(clus.labels)])
qc <- qc[names(clus.labels),]

vars <- rowVars(log1p(counts))
names(vars) <- rownames(counts)
vars <- sort(vars, decreasing = TRUE)
core <- counts[names(vars)[1:1000],]
```

We have `r ncol(counts)` cells.

@@

@@

# 0. Residuals for visualization
## Fit zinbwave X intercept, no batch
Let's run zinbwave with K = 0 and X batches.
```{r}
fn = '../data/zinb_k0_batch.rda'
if (runZinb){
  mod = model.matrix( ~ batch)
  print(system.time(zinb0 <- zinbFit(core, K = 0, X = mod, ncores = NCORES)))
  save(zinb0, file = fn)
}else{
  load(fn)
}
```

We use deviance residuals for visualization. Let's check that deviance residuals look ok.
```{r}
res = compute.deviance.residuals(core, zinb0)
res[1:3,1:3]
```

PCA on the residuals where color are for batches on the left and previously found clusters on the right.
```{r}
pca = prcomp(t(res))
par(mfrow = c(1,2))
plot(pca$x, col = batch, pch = 19,
     main="PCA of residuals\ncolor=batch")
plot(pca$x, col = clus.labels, pch = 19,
     main = "PCA of residuals\ncolor=cluster")
par(mfrow = c(1,1))
```

Boxplot of the residuals for each cell.
```{r}
res_order = res[, order(as.numeric(batch))]
col_order = as.numeric(batch)[order(as.numeric(batch))]
boxplot(res_order, main='Boxplot of residuals\ncolor=batch', col = col_order,         staplewex = 0, outline = 0, border = col_order, xaxt = 'n')
```

@@

Let's run zinbwave with K = 50 and X including batches.
```{r}
fn = '../data/zinb_k50_batch.rda'
if (runZinb){
  mod = model.matrix( ~ batch)
  print(system.time(zinb50 <- zinbFit(core, ncores = NCORES, K = 50,
                                      X = mod)))
  save(zinb50, file = fn)
}else{
  load(fn)
}
```

@@

```{r rsec_50, eval=FALSE}
fn = '../data/clustExp_W_k50.rda'
if(runClus) {
ceObj <- RSEC(t(W), k0s = 4:15, alphas = c(0.1), betas = 0.8,
                      clusterFunction = "hierarchical01", minSizes=1,
                      sequential = TRUE, subsample = TRUE, ncores = NCORES,
					  combineProportion = 0.7,mergeMethod = "locfdr", mergeCutoff = 0.01
                      subsampleArgs = list(resamp.num=100,
                                           clusterFunction="kmeans",
                                           clusterArgs=list(nstart=10)),
                      seqArgs = list(k.min=3, top.can=5), verbose=TRUE)
save(ceObj, file = fn)
}
else{
	load(fn)
}
```

```{r examineCombineMany, eval=FALSE}
plotClusters(ceObj, colPalette = c(bigPalette, rainbow(100)))
plotCoClustering(ceObj)
table(primaryClusterNamed(ceObj))
sum(primaryCluster(ceObj) == -1)
```

```{r examineMergeClusters, eval=FALSE}
#re-does merging simpling to make plot 
mergeClusters(ceObj, mergeMethod = "locfdr", plot="mergeMethod",cutoff = 0.01)
```

```{r heatmap,eval=FALSE}
plotHeatmap(ceObj, clusterSamplesData="dendrogramValue", breaks = .99)
```

@@

```{r consensus_50}
## matching the parameters used by Russell
W = zinb50@W
fn = '../data/clustExp_W_k50.rda'
if(runClus) {
  cl_res <- clusterMany(t(W), ks = 4:15, alphas = c(0.1), betas = 0.8,
                        clusterFunction = "hierarchical01", minSizes=1,
                        sequential = TRUE, subsample = TRUE, ncores = 7,
                        subsampleArgs = list(resamp.num=100,
                                             clusterFunction="kmeans",
                                             clusterArgs=list(nstart=10)),
                        seqArgs = list(k.min=3, top.can=5), verbose=TRUE)
  save(cl_res, file = fn)
} else {
  load(fn)
}
```

```{r combine_50}
combined <- combineMany(cl_res, proportion = 0.7)
plotClusters(combined, colPalette = c(bigPalette, rainbow(100)))
plotCoClustering(combined)
table(primaryClusterNamed(combined))
sum(primaryCluster(combined) == -1)
```

```{r new_clusters}
plot(getW(zinb50), col=clus.labels, main = 'color = previous')
plot(getW(zinb50), col=as.factor(primaryCluster(combined)), main = 'color = new')
table(data.frame(previous = clus.labels, new = primaryCluster(combined)))
```

```{r merged}
combined <- makeDendrogram(combined)
plotDendrogram(combined)
merged <- mergeClusters(combined, mergeMethod = "locfdr", cutoff = 0.01)
```

@@

```{r heatmap}
plotHeatmap(merged, clusterSamplesData="dendrogramValue", breaks = .99)
```

So, let's look at a heatmap on deviance residuals using heatmap.2 from gplots package with all the 1000 most variable genes.
```{r}
palDF = merged@clusterLegend[[1]]
pal = palDF[, 'color']
names(pal) = palDF[, 'name']

>>>>

```{r}
runZinb = TRUE
runClus = TRUE
NCORES = 2
```

@@

EAP: small request. Can everyone put a line between the beginning of a r chunk and text? It makes it nicely formated for my text editor.

@@

EAP: I have updated the code to pull from a dataset on the repos that is created with the `createData.R` file. For now, I am filtering to the top 1000 most variable genes there, though we might want to add that to the code for the article. This will be slightly different data from Russell's, which didn't use all of the samples. We can adjust that decision later, or just compare the samples that are the same. (Russell's clusterLabels are in the meta data)

```{r datain}
#counts<-read.table("../data/oeCufflinkCountData.txt",sep="\t",header=TRUE)
core<-read.table("../data/oeCufflinkCountData_1000Var.txt",sep="\t",header=TRUE)
metadata<-read.table("../data/oeMetadata.txt",sep="\t",header=TRUE)
# load("../data/Expt4c2b_filtdata.Rda")
# load("../data/E4c2b_slingshot_wsforkelly.RData")
```

Here we only look at the 1000 most variable genes. EAP: see note above, I've commented out the filtering and added it to the `createData.R` for now.

```{r filtering}
batch <- metadata$Batch
qc <- metadata[,!names(metadata)%in% c("Batch","Experiment","clusterLabels")]
clus.labels<-metadata[,"clusterLabels"]
# vars <- rowVars(log1p(counts))
# names(vars) <- rownames(counts)
# vars <- sort(vars, decreasing = TRUE)
# core <- counts[names(vars)[1:1000],]
```

We have `r ncol(core)` cells.

@@

EAP: if we are going to ultimately use these functions, it would be good to add them to the package for zinbwave.

@@

# 0. Residuals for visualization
## Fit zinbwave X intercept, no batch
Let's run zinbwave with K = 0 and X batches.

```{r}
fn = '../data/zinb_k0_batch.rda'
if (runZinb){
  mod = model.matrix( ~ batch)
  print(system.time(zinb0 <- zinbFit(core, K = 0, X = mod, ncores = NCORES)))
  save(zinb0, file = fn)
}else{
  load(fn)
}
```

We use deviance residuals for visualization. Let's check that deviance residuals look ok.

```{r}
res = compute.deviance.residuals(core, zinb0)
res[1:3,1:3]
```

PCA on the residuals where color are for batches on the left and previously found clusters on the right.

```{r}
pca = prcomp(t(res))
par(mfrow = c(1,2))
plot(pca$x, col = batch, pch = 19,
     main="PCA of residuals\ncolor=batch")
plot(pca$x, col = clus.labels, pch = 19,
     main = "PCA of residuals\ncolor=cluster")
par(mfrow = c(1,1))
```

Boxplot of the residuals for each cell.

```{r}
res_order = res[, order(as.numeric(batch))]
col_order = as.numeric(batch)[order(as.numeric(batch))]
boxplot(res_order, main='Boxplot of residuals\ncolor=batch', col = col_order,         staplewex = 0, outline = 0, border = col_order, xaxt = 'n')
```

@@

Let's run zinbwave with K = 50 and X including batches.

```{r}
fn = '../data/zinb_k50_batch.rda'
if (runZinb){
  mod = model.matrix( ~ batch)
  print(system.time(zinb50 <- zinbFit(core, ncores = NCORES, K = 50,
                                      X = mod)))
  save(zinb50, file = fn)
}else{
  load(fn)
}
```

@@

```{r rsec_50}
fn = '../data/clustExp_W_k50.rda'
if(runClus) {
ceObj <- RSEC(t(W), k0s = 4:15, alphas = c(0.1), betas = 0.8,
                      clusterFunction = "hierarchical01", minSizes=1,
                      sequential = TRUE, subsample = TRUE, ncores = NCORES,
					  combineProportion = 0.7,mergeMethod = "locfdr", mergeCutoff = 0.01
                      subsampleArgs = list(resamp.num=100,
                                           clusterFunction="kmeans",
                                           clusterArgs=list(nstart=10)),
                      seqArgs = list(k.min=3, top.can=5), verbose=TRUE)
save(ceObj, file = fn)
}
else{
	load(fn)
}
```

```{r examineCombineMany}
plotClusters(ceObj, colPalette = c(bigPalette, rainbow(100)))
plotCoClustering(ceObj)
table(primaryClusterNamed(ceObj))
sum(primaryCluster(ceObj) == -1)
```

```{r examineMergeClusters}
#re-does merging simpling to make plot 
if(packageVersion("clusterExperiment")>='1.3.0.9001') mergeClusters(ceObj, mergeMethod = "locfdr", plotInfo="mergeMethod",cutoff = 0.01)
```

```{r heatmap }
if(packageVersion("clusterExperiment")>='1.3.0.9002') plotHeatmap(ceObj, clusterSamplesData="dendrogramValue", breaks = .99)
```

```{r compareClusters}
plot(getW(zinb50), col=clus.labels, main = 'color = previous')
plot(getW(zinb50), col=as.factor(primaryCluster(ceObj)), main = 'color = new')
table(data.frame(previous = clus.labels, new = primaryCluster(ceObj)))
```

@@

```{r consensus_50, eval=FALSE}
## matching the parameters used by Russell
W = zinb50@W
fn = '../data/clustExp_W_k50.rda'
if(runClus) {
  cl_res <- clusterMany(t(W), ks = 4:15, alphas = c(0.1), betas = 0.8,
                        clusterFunction = "hierarchical01", minSizes=1,
                        sequential = TRUE, subsample = TRUE, ncores = 7,
                        subsampleArgs = list(resamp.num=100,
                                             clusterFunction="kmeans",
                                             clusterArgs=list(nstart=10)),
                        seqArgs = list(k.min=3, top.can=5), verbose=TRUE)
  save(cl_res, file = fn)
} else {
  load(fn)
}
```

```{r combine_50, eval=FALSE}
combined <- combineMany(cl_res, proportion = 0.7)
plotClusters(combined, colPalette = c(bigPalette, rainbow(100)))
plotCoClustering(combined)
table(primaryClusterNamed(combined))
sum(primaryCluster(combined) == -1)
```

```{r new_clusters, eval=FALSE}
plot(getW(zinb50), col=clus.labels, main = 'color = previous')
plot(getW(zinb50), col=as.factor(primaryCluster(combined)), main = 'color = new')
table(data.frame(previous = clus.labels, new = primaryCluster(combined)))
```

```{r merged, eval=FALSE}
combined <- makeDendrogram(combined)
plotDendrogram(combined)
merged <- mergeClusters(combined, mergeMethod = "locfdr", cutoff = 0.01)
```

@@

```{r heatmapOld, eval=FALSE}
plotHeatmap(merged, clusterSamplesData="dendrogramValue", breaks = .99)
```

So, let's look at a heatmap on deviance residuals using heatmap.2 from gplots package with all the 1000 most variable genes.
```{r, eval=FALSE}
palDF = merged@clusterLegend[[1]]
pal = palDF[, 'color']
names(pal) = palDF[, 'name']
